---
title: "QC_Microbiome_Alpha_Beta_RelativeAbundance"
author: "Deepa Upreti"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# 1. Load packages
library(phyloseq)
library(ggplot2)
library(dplyr)
library(vegan)
library(RColorBrewer)
library(openxlsx)
library(readr)
library(plotly)
library(viridisLite)
library(htmlwidgets)
library(patchwork)
library(scales)


# 2. Import data 
#Input files
otu_file  <- "taxatable-filtered-absolute_w_taxonomy.txt" 
meta_file <- "metadata.tsv"

# Read taxonomy/count table
# Taxonomy strings are stored in the first column/row names
taxatable <- read.table(
  otu_file,
  header = TRUE,
  sep = "\t",
  row.names = 1,
  check.names = FALSE
)

# Inspect column names
head(colnames(taxatable))

#  Remove the "OTU ID" column (second column)
taxatable <- taxatable[, -1]

# 3. Separate count and taxonomy information
#  Extract counts and taxonomy
otu_counts <- taxatable[, sapply(taxatable, is.numeric)]  # Counts only
taxonomy_strings <- rownames(taxatable)         # Taxonomy strings are rownames

#  Split taxonomy into hierarchical levels
tax_split <- data.frame(do.call(rbind, strsplit(taxonomy_strings, ";\\s*")))
colnames(tax_split) <- c("Kingdom","Phylum","Class","Order","Family","Genus","Species","Strain")[1:ncol(tax_split)]
rownames(tax_split) <- rownames(otu_counts)

#4. Import and align sample metadata
meta <- read.table(
  meta_file,
  header = TRUE,
  row.names = 1,
  sep = "\t",
  check.names = FALSE
)
head(meta)

# Check for unmatched sample IDs
setdiff(colnames(otu_counts), rownames(meta))
setdiff(rownames(meta), colnames(otu_counts))

common_samples <- intersect(colnames(otu_counts), rownames(meta))

otu_counts <- otu_counts[, common_samples, drop = FALSE]
meta <- meta[common_samples, , drop = FALSE]

stopifnot(identical(colnames(otu_counts), rownames(meta)))
```

```{r}
#5. Build phyloseq object
ps <- phyloseq(
  otu_table(as.matrix(otu_counts), taxa_are_rows = TRUE),
  tax_table(as.matrix(tax_split)),
  sample_data(meta)
)

# Quick check
ps

#Quick integrity check for OTU–metadata alignment

cat("Number of taxa:", ntaxa(ps), "\nNumber of samples:", nsamples(ps), "\n\n")

# Show any sample names that don't match (should be empty)
setdiff(colnames(otu_table(ps)), rownames(sample_data(ps)))

# Confirm sample sums (none should be 0)
summary(sample_sums(ps))
# Confirm OTU and metadata sample IDs match
identical(sort(colnames(otu_table(ps))), sort(rownames(sample_data(ps))))
```

```{r}
# ============================================
# Rename exposure groups
# ============================================

sample_data(ps)$SampleType <- dplyr::recode(
  as.character(sample_data(ps)$SampleType),
  "Control" = "CON",
  "PAE"     = "ALC",
  "CBD"     = "CB",
  "SAC"     = "ALC+CB"
)

sample_data(ps)$SampleType <- factor(
  sample_data(ps)$SampleType,
  levels = c("CON", "ALC", "CB", "ALC+CB")
)

# Verify group labels and sample numbers
table(sample_data(ps)$SampleType)
```


Sequencing depth on the basis of Treatment
```{r}
depth_df <- data.frame(
  Sample = sample_names(ps),
  Reads  = sample_sums(ps),
  Group  = sample_data(ps)$SampleType  # change 'Group' to your metadata variable name
)

ggplot(depth_df, aes(x = Group, y = Reads, fill = Group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6) +
  theme_bw(base_size = 13) +
  ylab("Sequencing depth (reads)") +
  ggtitle("Library size per sample")

summary(sample_sums(ps))
# Check variability ratio
cat("Max/Min read-depth ratio:", max(sample_sums(ps)) / min(sample_sums(ps)), "\n")


# Sort by total reads
depth_df_sorted <- depth_df[order(depth_df$Reads, decreasing = TRUE), ]
head(depth_df_sorted, 5)   # top 5 most-sequenced
tail(depth_df_sorted, 5)   # bottom 5 least-sequenced

ggplot(depth_df, aes(x = "", y = Reads)) +
  geom_boxplot(outlier.shape = NA, fill = "skyblue") +
  geom_jitter(width = 0.1, alpha = 0.6) +
  scale_y_log10() +
  theme_bw(base_size = 13) +
  ylab("Reads (log10)") +
  ggtitle("Sequencing depth per sample")

```

Outlier detection by sequencing depth
```{r}

# Create a data frame of sample depths
depth_df <- data.frame(
  Sample = sample_names(ps),
  Reads  = sample_sums(ps),
  Group  = sample_data(ps)$SampleType  # change to your grouping variable
)

# Calculate mean and standard deviation
mean_depth <- mean(depth_df$Reads)
sd_depth   <- sd(depth_df$Reads)


# Flag samples with sequencing depth > mean + 2 SD
depth_df$Outlier <- depth_df$Reads > (mean_depth + 2 * sd_depth)


# Show flagged samples
cat("Samples flagged by sequencing-depth criterion (>2 SD above mean):\n")
print(
  depth_df[
    depth_df$Outlier,
    c("Sample", "Reads", "Group")
  ]
)

# Store flagged sample IDs for QC review
outlier_samples <- depth_df$Sample[depth_df$Outlier]
outlier_samples
```


# Additional sample-level QC diagnostics
```{r}
# Bray-Curtis distance-based outlier detection
# Identifies samples that are unusually dissimilar from the overall dataset.

ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
bray <- phyloseq::distance(ps_rel, method = "bray")

# Average distance of each sample to all others
bray_mat <- as.matrix(bray)
avg_dist <- rowMeans(bray_mat)

# Convert to z-scores
z_dist <- scale(avg_dist)

# Flag samples > 3 SD
outliers_dist <- names(z_dist[abs(z_dist) > 3])
outliers_dist

# Ordination-based outlier detection
ord <- ordinate(ps_rel, method="PCoA", distance=bray)

coords <- ord$vectors[,1:3]   # first 3 axes
centroid <- colMeans(coords)
dist_centroid <- sqrt(rowSums((coords - centroid)^2))

z_ord <- scale(dist_centroid)

outliers_ord <- names(z_ord[abs(z_ord) > 3])
outliers_ord


#Alpha diversity outlier detection
# Extreme richness/evenness indicates artifact.
alpha <- estimate_richness(ps, measures=c("Observed","Shannon"))
alpha$z_shannon <- scale(alpha$Shannon)
alpha$z_observed <- scale(alpha$Observed)

alpha_outliers <- rownames(alpha)[
  abs(alpha$z_shannon) > 3 | abs(alpha$z_observed) > 3
]
alpha_outliers

#Detect if one taxon dominates abnormally (Dominance / contamination signal)
otu <- as(otu_table(ps), "matrix")
if (!taxa_are_rows(ps)) otu <- t(otu)

rel <- sweep(otu, 2, colSums(otu), "/")
max_rel <- apply(rel, 2, max)

z_dom <- scale(max_rel)
outliers_dom <- names(z_dom[abs(z_dom) > 3])
outliers_dom


#COMBINE QC results (Summary)
outlier_summary <- data.frame(
  Sample = sample_names(ps),
  BrayOutlier = sample_names(ps) %in% outliers_dist,
  OrdOutlier  = sample_names(ps) %in% outliers_ord,
  AlphaOutlier = sample_names(ps) %in% alpha_outliers,
  DominanceOutlier = sample_names(ps) %in% outliers_dom
)

outlier_summary$TotalFlags <- rowSums(outlier_summary[,2:5])
outlier_summary[order(outlier_summary$TotalFlags, decreasing=TRUE),]

###NOTE: One sample exhibited extreme sequencing depth (~23 million reads), exceeding the next highest sample by more than fivefold and resulting in substantial library size imbalance (Max/Min = 27.5; CV = 1.23). To maintain comparable sequencing depth across samples and minimize technical bias, this sample was excluded prior to downstream diversity and differential abundance analyses. Key findings were consistent when analyses were performed with and without this sample.
#| Metric                        | Full dataset (n=65) | After removal (n=64) |
#| ----------------------------- | ------------------- | -------------------- |
#| Minimum reads                 | 856,108             | 856,108              |
#| Median reads                  | 1,815,354           | 1,803,071            |
#| Mean reads                    | 2,278,197           | 1,945,608            |
#| Maximum reads                 | **23,563,850**      | 4,292,139            |
#| Max / Min ratio               | **27.5**            | 5.01                 |
#| Coefficient of variation (CV) | **1.23**            | 0.41                 |

```


```{r}
# ============================================
# Final sample exclusion
# ============================================

# MICE.SEQ13 was excluded from downstream analyses because of
# extreme sequencing depth and substantial library-size imbalance.

ps_filtered <- prune_samples(
  sample_names(ps) != "MICE.SEQ13",
  ps
)

# Confirm that the intended sample was removed
stopifnot(!("MICE.SEQ13" %in% sample_names(ps_filtered)))

# Confirm that exactly one sample was removed
stopifnot(nsamples(ps_filtered) == nsamples(ps) - 1)

cat(
  "Samples before QC exclusion:", nsamples(ps),
  "\nSamples after QC exclusion:", nsamples(ps_filtered),
  "\n"
)
```



ALPHA and BETA DIVERSITY  (Sample no.13 removed)

#ALPHA DIVERSITY
```{r}

alpha_beta_dir <- "Microbiome_Alpha_Beta_Output"

if (!dir.exists(alpha_beta_dir)) {
  dir.create(alpha_beta_dir)
}

# ============================================
# Alpha diversity
# MICE.SEQ13 excluded based on sequencing-depth QC
# Alpha-diversity metrics are calculated from raw counts.
# ============================================
ps_alpha<-ps_filtered


# Compute alpha-diversity indices
alpha_df <- estimate_richness(
  ps_alpha,
  measures = c("Observed", "Chao1", "Shannon", "Simpson")
)

# Add sample metadata
alpha_df$SampleID   <- rownames(alpha_df)
alpha_df$SampleType <- as.vector(sample_data(ps_alpha)$SampleType)

alpha_df$Sex <- if ("Sex" %in% colnames(sample_data(ps_alpha))) {
  as.vector(sample_data(ps_alpha)$Sex)
} else {
  NA
}
###Quality check if the samples are in same order
nsamples(ps_alpha)
nrow(alpha_df)

identical(sample_names(ps_alpha), alpha_df$SampleID)

all(alpha_df$SampleID %in% sample_names(ps_alpha))
"MICE.SEQ13" %in% sample_names(ps_alpha) #check sample is removed; should come False


# Reorder columns (GraphPad-friendly)
alpha_df <- alpha_df[, c(
  "SampleID", "SampleType", "Sex",
  "Observed", "Chao1", "Shannon", "Simpson"
)]

# Set desired order
alpha_df$SampleType <- factor(alpha_df$SampleType, levels = c("CON","ALC","CB","ALC+CB"))
alpha_df$Sex <- factor(alpha_df$Sex, levels = c("M","F"))
alpha_df <- alpha_df[order(alpha_df$SampleType, alpha_df$Sex), ]

# Sanity check (should show 4 groups)
table(alpha_df$SampleType, useNA="ifany")


# Export for GraphPad Prism
write.csv(
  alpha_df,
  file.path(alpha_beta_dir, "Alpha_Diversity_Raw_counts_no13.csv"),
  row.names = FALSE
)

summary(sample_sums(ps_alpha))


# ---- Alpha diversity plots ----

p_shannon <- ggplot(alpha_df, aes(x = SampleType, y = Shannon, fill = SampleType)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
  theme_bw(base_size = 13) +
  ylab("Shannon diversity index") +
  xlab("") +
  ggtitle("Alpha diversity (Shannon)") +
  theme(legend.position = "none")

p_obs <- ggplot(alpha_df, aes(x = SampleType, y = Observed, fill = SampleType)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
  theme_bw(base_size = 13) +
  ylab("Observed richness") +
  xlab("") +
  ggtitle("Alpha diversity (Observed)") +
  theme(legend.position = "none")

p_chao1 <- ggplot(alpha_df, aes(x = SampleType, y = Chao1, fill = SampleType)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
  theme_bw(base_size = 13) +
  ylab("Chao1 richness estimate") +
  xlab("") +
  ggtitle("Alpha diversity (Chao1)") +
  theme(legend.position = "none")

p_alpha_all <- p_obs | p_chao1 | p_shannon
p_alpha_all

ggsave(
  file.path(alpha_beta_dir, "Alpha_Diversity_Observed_Chao1_Shannon_no13.pdf"),
  plot     = p_alpha_all,
  width    = 12,
  height   = 4,
  units    = "in"
)

ggsave(
  file.path(alpha_beta_dir, "Alpha_Shannon_no13.pdf"),
  p_shannon, width = 5, height = 4
)
ggsave(
  file.path(alpha_beta_dir, "Alpha_Observed_no13.pdf"),
  p_obs, width = 5, height = 4
)
ggsave(
  file.path(alpha_beta_dir, "Alpha_Chao1_no13.pdf"),
  p_chao1, width = 5, height = 4
)

table(alpha_df$SampleType)


###Quality check
# 1) sample removed?
"MICE.SEQ13" %in% sample_names(ps_alpha)     # should be FALSE

# 2) group counts look reasonable (not all NA, not 1 group)
table(alpha_df$SampleType, useNA="ifany")    # should show Control/PAE/CBD/SAC

# 3) Sex counts okay
table(alpha_df$Sex, useNA="ifany")           # should show M/F (and ideally no NAs)

```

# Sequencing-depth QC summary
```{r}

depth_qc_summary <- function(ps_obj) {
  d <- sample_sums(ps_obj)
  c(
    n_samples = nsamples(ps_obj),
    min_reads = min(d),
    median_reads = median(d),
    mean_reads = mean(d),
    max_reads = max(d),
    max_min_ratio = max(d) / min(d),
    cv = sd(d) / mean(d)
  )
}

qc_allsamples <- depth_qc_summary(ps)
qc_removeno13 <- depth_qc_summary(ps_filtered)

qc_table <- data.frame(
  Metric = names(qc_allsamples),
 All_samples = qc_allsamples,
  Removed_no13sample = qc_removeno13 
)

qc_table

###Table format
qc_table$All_samples  <- round(qc_table$All_samples, 2)
qc_table$Removed_no13sample <- round(qc_table$Removed_no13sample, 2)

qc_table

write.csv(qc_table,
          file.path(alpha_beta_dir, "SequencingDepth_QC_table_no13.csv"),
          row.names = FALSE)
```

Sequencing depth outlier plot

```{r}
depth_df <- data.frame(
  Sample = sample_names(ps),
  Reads  = sample_sums(ps),
  Sex    = sample_data(ps)$Sex
)

depth_df$Outlier <- ifelse(depth_df$Sample == "MICE.SEQ13", "Outlier", "Other")

p_depth <- ggplot(depth_df, aes(x = "", y = Reads)) +
  geom_boxplot(outlier.shape = NA, fill = "grey85") +
  geom_jitter(aes(color = Outlier), width = 0.08, size = 3) +
  scale_color_manual(values = c("Outlier" = "red", "Other" = "black")) +
  scale_y_log10() +
  theme_bw(base_size = 14) +
  ylab("Sequencing depth (reads, log10 scale)") +
  xlab("") +
  ggtitle("Sequencing depth across samples") +
  theme(
    legend.title = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

# Save figure
ggsave(file.path(alpha_beta_dir,
  "Sequencing_depth_outlier_plot.pdf"),
  plot = p_depth,
  width = 6,
  height = 5
)

p_sex_depth <- ggplot(depth_df, aes(x = Sex, y = Reads, fill = Sex)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(aes(color = Outlier), width = 0.15, size = 3) +
  scale_color_manual(values = c("Outlier" = "red", "Other" = "black")) +
  scale_y_log10() +
  theme_bw(base_size = 14) +
  ylab("Sequencing depth (reads, log10 scale)") +
  ggtitle("Sequencing depth by sex") +
  theme(
    legend.title = element_blank()
  )


# Save figure
ggsave(file.path(alpha_beta_dir,
  "Sequencing_depth_outlier_bySex.pdf"),
  plot = p_sex_depth,
  width = 6,
  height = 5
)
p_sex_depth
```

```{r}
# ============================================
# Beta Diversity (Bray–Curtis, sample 13 removed)
# ============================================
# --------------------------------------------
# Use filtered dataset (sample 13 removed)
# --------------------------------------------
ps_beta <- ps_filtered

# Verify removal
"MICE.SEQ13" %in% sample_names(ps_beta)

# --------------------------------------------
# Normalize to relative abundance
# --------------------------------------------
ps_rel <- transform_sample_counts(ps_beta, function(x) x / sum(x))

# --------------------------------------------
# Bray–Curtis distance
# --------------------------------------------
bray_dist <- phyloseq::distance(ps_rel, method = "bray")

# --------------------------------------------
# PCoA ordination
# --------------------------------------------
ord <- ordinate(ps_rel, method = "PCoA", distance = bray_dist)

# Percent variance explained
eigvals <- ord$values$Eigenvalues
var_exp <- round(100 * eigvals / sum(eigvals), 2)

# --------------------------------------------
# 2D PCoA plot
# --------------------------------------------
p_beta2D <- plot_ordination(ps_rel, ord, color = "SampleType") +
  geom_point(size = 3, alpha = 0.85) +
  stat_ellipse(aes(color = SampleType),
               type = "t",
               linetype = 2,
               linewidth = 1) +
  labs(
    x = paste0("PCoA1 (", var_exp[1], "%)"),
    y = paste0("PCoA2 (", var_exp[2], "%)")
  ) +
  theme_bw(base_size = 13) +
  ggtitle("Beta diversity (Bray–Curtis)") +
  theme(legend.position = "right")

p_beta2D

# Save 2D PCoA
ggsave(
  file.path(alpha_beta_dir,"Beta_Diversity_PCoA_BrayCurtis_no13.pdf"),
  plot = p_beta2D,
  width = 6,
  height = 5
)

# --------------------------------------------
# PERMANOVA
# --------------------------------------------
meta_df <- data.frame(sample_data(ps_rel))

set.seed(123)

perm <- adonis2(
  bray_dist ~ Sex + SampleType,
  data = meta_df,
  permutations = 999,
  by = "terms"
)

print(perm)

# --------------------------------------------
# Dispersion test
# --------------------------------------------
disp <- betadisper(bray_dist, meta_df$SampleType)

anova(disp)
permutest(disp, permutations = 999)

# --------------------------------------------
# Combine Alpha + Beta
# --------------------------------------------
p_combined <- (p_alpha_all | p_beta2D) +
  plot_annotation(title = "Alpha and Beta Diversity Overview (Sample 13 removed)")

p_combined

ggsave(
  file.path(alpha_beta_dir,"Alpha_Beta_Diversity_Plots_no13.pdf"),
  plot = p_combined,
  width = 12,
  height = 5
)

# --------------------------------------------
# Save Bray–Curtis matrix
# --------------------------------------------
write.csv(
  as.matrix(bray_dist),
  file.path(alpha_beta_dir, "BrayCurtis_Distance_Matrix.csv")
)

# ============================================
# 3D PCoA Visualization
# ============================================

ord_df <- data.frame(
  Axis1 = ord$vectors[,1],
  Axis2 = ord$vectors[,2],
  Axis3 = ord$vectors[,3],
  SampleType = meta_df$SampleType,
  SampleID = rownames(meta_df)
)

cols <- viridisLite::plasma(length(unique(ord_df$SampleType)))

p_beta3D <- plot_ly(
  data = ord_df,
  x = ~Axis1,
  y = ~Axis2,
  z = ~Axis3,
  color = ~SampleType,
  colors = cols,
  type = "scatter3d",
  mode = "markers",
  text = ~SampleID,
  marker = list(size = 6, opacity = 0.85)
) %>%
  layout(
    title = "3D PCoA (Bray–Curtis)",
    scene = list(
      xaxis = list(title = paste0("PCoA1 (", var_exp[1], "%)")),
      yaxis = list(title = paste0("PCoA2 (", var_exp[2], "%)")),
      zaxis = list(title = paste0("PCoA3 (", var_exp[3], "%)"))
    )
  )

p_beta3D

saveWidget(
  p_beta3D,
  file.path(alpha_beta_dir, "Beta_PCoA_3D.html"),
  selfcontained = TRUE
)
```

Sex-based Bray–Curtis PCoA (2D)
```{r}
# ============================================
# Beta Diversity by Sex
# ============================================

# Use same normalized dataset
ps_rel_sex <- ps_rel

# Bray–Curtis distance
bray_sex <- phyloseq::distance(ps_rel_sex, method = "bray")

# PCoA ordination
ord_sex <- ordinate(ps_rel_sex, method = "PCoA", distance = bray_sex)

# Percent variance explained
eigvals_sex <- ord_sex$values$Eigenvalues
var_exp_sex <- round(100 * eigvals_sex / sum(eigvals_sex), 2)

# Plot
p_beta_sex <- plot_ordination(ps_rel_sex, ord_sex,
                              color = "Sex") +
  geom_point(size = 3, alpha = 0.85) +
  stat_ellipse(aes(color = Sex),
               type = "t",
               linetype = 2,
               linewidth = 1) +
  labs(
    x = paste0("PCoA1 (", var_exp_sex[1], "%)"),
    y = paste0("PCoA2 (", var_exp_sex[2], "%)")
  ) +
  theme_bw(base_size = 13) +
  ggtitle("Beta diversity by Sex (Bray–Curtis)") +
  theme(legend.position = "right")

p_beta_sex

# Save figure
ggsave(
  file.path(alpha_beta_dir,"Beta_Diversity_PCoA_Sex_BrayCurtis_no13.pdf"),
  plot = p_beta_sex,
  width = 6,
  height = 5
)
```

Separate PCA plot (Pairwise bray-curtis and Permanova and include sex)
```{r}

# Pairwise PCoA function with q-value support
pairwise_pcoa_plot <- function(ps_obj, group1, group2, q_val = NA){

  meta_full <- data.frame(sample_data(ps_obj))

  keep_samples <- rownames(meta_full)[
    meta_full$SampleType %in% c(group1, group2)
  ]

  ps_sub <- prune_samples(keep_samples, ps_obj)

  ps_rel <- transform_sample_counts(ps_sub, function(x) x / sum(x))

  bray <- phyloseq::distance(ps_rel, method = "bray")

  ord <- ordinate(ps_rel, method = "PCoA", distance = bray)

  eigvals <- ord$values$Eigenvalues
  var_exp <- round(100 * eigvals / sum(eigvals), 2)

  meta <- data.frame(sample_data(ps_rel))

  perm <- adonis2(
  bray ~ Sex + SampleType,
  data = meta,
  permutations = 999,
  by = "terms"
)
  R2_val <- perm["SampleType", "R2"]
p_val  <- perm["SampleType", "Pr(>F)"]

# PERMANOVA uses permutation testing to assess differences in
# multivariate community composition based on Bray-Curtis distances.

  if(!is.na(q_val)){
    label_text <- paste0(
      "R² = ", sprintf("%.3f", R2_val),
      "\nq = ", sprintf("%.4f", q_val)
    )
  } else {
    label_text <- paste0(
      "R² = ", sprintf("%.3f", R2_val),
      "\np = ", signif(p_val, 3)
    )
  }

  ord_df <- data.frame(
    Axis1 = ord$vectors[,1],
    Axis2 = ord$vectors[,2],
    SampleType = meta$SampleType,
    Sex = meta$Sex
  )

  p <- ggplot(ord_df, aes(Axis1, Axis2, color = SampleType, shape = Sex)) +
    geom_point(size = 3, alpha = 0.9) +
    stat_ellipse(
      aes(group = SampleType, color = SampleType),
      type = "t",
      linewidth = 1,
      show.legend = FALSE
    ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +

    annotate(
  "label",
  x = 0.65,
  y = -0.45,
  label = label_text,
  hjust = 1,
  vjust = 0,
  size = 3.8,
  fill = "white",
  alpha = 0.9
) +

    scale_color_manual(values = c(
  "CON"    = "#1f78b4",
  "ALC"    = "#e31a1c",
  "CB"     = "#33a02c",
  "ALC+CB" = "#6a3d9a"
)) +

    labs(
      x = paste0("PCoA1 (", var_exp[1], "%)"),
      y = paste0("PCoA2 (", var_exp[2], "%)"),
      color = "Treatment",
      shape = "Sex"
    ) +

    ggtitle(paste(group1, "vs", group2)) +

    coord_cartesian(clip = "off") +

    theme_bw(base_size = 12) +

    theme(
      panel.grid = element_blank(),
      plot.title = element_text(
        hjust = 0.5,
        margin = margin(b = 15)
      ),
      plot.margin = margin(10, 10, 35, 10)
    )

  return(list(
    plot = p,
    R2 = R2_val,
    pvalue = p_val
  ))
}

# Run pairwise comparisons
ps_beta <- ps_filtered

# Set seed for reproducible permutation tests
set.seed(123)
c1 <- pairwise_pcoa_plot(ps_beta, "CON", "ALC")
c2 <- pairwise_pcoa_plot(ps_beta, "CON", "CB")
c3 <- pairwise_pcoa_plot(ps_beta, "CON", "ALC+CB")
c4 <- pairwise_pcoa_plot(ps_beta, "CB", "ALC+CB")
c5 <- pairwise_pcoa_plot(ps_beta, "ALC", "CB")
c6 <- pairwise_pcoa_plot(ps_beta, "ALC", "ALC+CB")

# Create PERMANOVA table
perm_table <- data.frame(
  Comparison = c(
    "CON vs ALC",
    "CON vs CB",
    "CON vs ALC+CB",
    "CB vs ALC+CB",
    "ALC vs CB",
    "ALC vs ALC+CB"
  ),
  R2 = c(c1$R2, c2$R2, c3$R2, c4$R2, c5$R2, c6$R2),
  pvalue = c(c1$pvalue, c2$pvalue, c3$pvalue, c4$pvalue, c5$pvalue, c6$pvalue)
)

# Benjamini-Hochberg correction
perm_table$qvalue <- p.adjust(perm_table$pvalue, method = "BH")

# Replot using q-values
set.seed(123)
c1 <- pairwise_pcoa_plot(ps_beta, "CON", "ALC", perm_table$qvalue[1])
c2 <- pairwise_pcoa_plot(ps_beta, "CON", "CB", perm_table$qvalue[2])
c3 <- pairwise_pcoa_plot(ps_beta, "CON", "ALC+CB", perm_table$qvalue[3])
c4 <- pairwise_pcoa_plot(ps_beta, "CB", "ALC+CB", perm_table$qvalue[4])
c5 <- pairwise_pcoa_plot(ps_beta, "ALC", "CB", perm_table$qvalue[5])
c6 <- pairwise_pcoa_plot(ps_beta, "ALC", "ALC+CB", perm_table$qvalue[6])

# Combine panels
p_panel <- (c1$plot | c2$plot | c3$plot) /
           (c4$plot| c5$plot | c6$plot)

p_panel

# Save figure
ggsave(
  file.path(alpha_beta_dir, "Pairwise_PCoA_panels_with_qvalues.pdf"),
  plot = p_panel,
  width = 16,
  height = 10
)

# Save table
write.csv(
  perm_table,
  file.path(alpha_beta_dir, "Pairwise_PERMANOVA_results.csv"),
  row.names = FALSE
)

perm_table
```

# Export relative-abundance tables for visualization and downstream summaries
# Counts are converted to relative abundance (0–1) after removal of MICE.SEQ13.
# No additional prevalence or abundance filtering is applied here.
```{r}
# Do NOT apply any prevalence filtering
#    We will use ps_filtered (remove outlier no13) exactly as is.

# ==============================
# Create output folder
# ==============================
output_dir <- "Microbiome_RelAbundance_Output"

if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}
# Convert counts → relative abundance (0–1)###TSS conversion
ps_rel <- transform_sample_counts(ps_filtered, function(x) x / sum(x)) ### remove the outlier sample no. 13


# Extract metadata
meta_df <- as.data.frame(sample_data(ps_rel))
meta_df$SampleID <- rownames(meta_df)

# Create Excel workbook
wb <- createWorkbook()

tax_ranks <- c("Kingdom","Phylum","Class","Order","Family","Genus","Species")

for (rank in tax_ranks) {

  if (!(rank %in% colnames(tax_table(ps_rel)))) next

  # Aggregate at taxonomic rank
  ps_rank <- tax_glom(ps_rel, taxrank = rank)

  mat <- as.data.frame(otu_table(ps_rank))
  if (!taxa_are_rows(ps_rank)) mat <- t(mat)

  tax_df <- as.data.frame(tax_table(ps_rank))

  # Replace NA/blank with "Unclassified"
  tax_names <- as.character(tax_df[[rank]])
  tax_names[is.na(tax_names) | tax_names == ""] <- "Unclassified"

  # Make unique names
  tax_names <- make.unique(tax_names)

  # Samples as rows
  mat2 <- as.data.frame(t(mat))
  colnames(mat2) <- tax_names
  mat2$SampleID <- rownames(mat2)

  # Join metadata
  final_df <- left_join(mat2, meta_df, by = "SampleID") %>%
    arrange(SampleType, Sex) %>%
    select(SampleID, SampleType, Sex, everything())

  addWorksheet(wb, rank)
  writeData(wb, rank, final_df)
}

saveWorkbook(
  wb,
  file.path(output_dir,"Relative_Abundance_All_Taxonomic_Levels_with_Metadata_nofilters_nosample13.xlsx"),
  overwrite = TRUE
)

cat("Export completed successfully (relative abundance 0–1, no filtering).\n")

```

```{r}
# Record R and package versions
sessionInfo()
```




